---
title: "Lan Vu HW5"
format: pdf
editor: visual
---

## Run the code and library

```{r}
library(readxl)
library(tidyverse)
library(psych)
```

```{r}
df <- read_excel("~/Downloads/DCCT.xlsx")
View(df)
```

## Part 1

1.  **Univariate Analysis**

```{r}
df %>%
  select(BMI12, BMI00, AGE) %>%
  psych::describe()


#BMI 12
ggplot(df, aes(x = BMI12)) +
  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Histogram of BMI at 12 Months", x = "BMI12", y = "Count")

ggplot(df, aes(y = BMI12)) +
  geom_boxplot(fill = "lightgreen") +
  theme_minimal() +
  labs(title = "Boxplot of BMI12", y = "BMI12")

#BMI10

ggplot(df, aes(x = BMI00)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "black") +
  theme_minimal() +
  labs(title = "Histogram of Baseline BMI", x = "BMI00", y = "Count")

ggplot(df, aes(y = BMI00)) +
  geom_boxplot(fill = "lightgreen") +
  theme_minimal() +
  labs(title = "Boxplot of BMI00", y = "BMI00")


#Age
ggplot(df, aes(x = AGE)) +
  geom_histogram(bins = 15, fill = "orange", color = "black") +
  theme_minimal() +
  labs(title = "Histogram of Age", x = "Age", y = "Count")

ggplot(df, aes(y = AGE)) +
  geom_boxplot(fill = "lightyellow") +
  theme_minimal() +
  labs(title = "Boxplot of AGE", y = "Age")



```

**AGE**

For variable age, we can see that it has the mean of 34.86 and median of 34.83 its value range from 21.00 to 48.00.

From the histogram we can see that the graph is somewhat normal and uniform even though there is no peak,

BMI AT 12 MONTHS

We can see that there are some missing values in this variable as its seems to have lower number of observations compare to the other 2 variables. It have the mean ad median at 25.14 and 24.75 respectively. From the graph, we can see that its distribution had a little bit of a right-skewed tails not too much. Overall it seems to have a normal distribution with some outliners in the far right of the graph.

BASELINE BMI

From the table, we can see that people participate in this trial has a mean BMI of 23.72 and median of 23.58. Theirs BMI score range from 16.28 to 34.39. The distriubtion of baseline BMI is quite normally distributed.

From the boxplots, we can see the that the two whisker are somewhat sysmestric. However, we can see some outliers which can be worth investigate later in the analysis phrase

Distributions of GROUP and GENDER

```{r}
table(df$GROUP)
prop.table(table(df$GROUP))

table(df$GENDER)
prop.table(table(df$GENDER))


#FREQ TABLE

ggplot(df, aes(x = GROUP, fill = GROUP)) +
  geom_bar(color = "black") +
  scale_fill_manual(values = c("INT" = "#1b9e77", "STD" = "#d95f02")) +
  labs(title = "Distribution of Treatment Group", x = "Treatment Group", y = "Count") +
  theme_minimal()


ggplot(df, aes(x = GENDER, fill = GENDER)) +
  geom_bar(color = "black") +
  scale_fill_manual(values = c("INT" = "#1b9e77", "STD" = "#d95f02")) +
  labs(title = "Distribution of Treatment Group", x = "GENDER", y = "Count") +
  theme_minimal()

```

For Groups variable:\
Based on the frequency table, we can tell that the disitribution within the groups is quite similar.

However, in Gender, we can tell that there are more male in this trial study compare to females participators.

2.  **Bivariate Analysis**

a\.

```{r}
# Boxplot of BMI12 by treatment group
ggplot(df, aes(x = GROUP, y = BMI12, fill = GROUP)) +
  geom_boxplot(color = "black") +
  scale_fill_manual(values = c("INT" = "#1b9e77", "STD" = "#d95f02")) +
  labs(title = "BMI at 12 Months by Treatment Group",
       x = "Treatment Group",
       y = "BMI12 (kg/m²)") +
  theme_minimal()

```

## Running Code

## Running Code

## Running Code

## Running Code Running Code

## Running Code

## Running Code

## Running Code
